---
title: "Lab One: Part Two"
editor: visual
---

---
# Introduction to R Lab

## Learning Objectives

By the end of this lab, you will be able to:

- Navigate the R environment and understand basic syntax
- Work with different data types and structures in R
- Import, explore, and manipulate datasets
- Create basic and intermediate visualizations
- Perform statistical analyses and interpret results
- Write custom functions and use control structures
---

## Part 1: R Fundamentals and Data Structures (60 minutes)

### Getting Started with R (10 minutes)

R is both a programming language and an environment for statistical computing. Let's start with the basics.

#### Basic Arithmetic Operations

```{r}
#| echo: true
#| eval: false
# R as a calculator
2 + 3
10 - 4
6 * 7
20 / 4
2^3  # exponentiation
sqrt(16)  # square root
log(10)   # natural logarithm
exp(1)    # e^1
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# R as a calculator
2 + 3
10 - 4
6 * 7
20 / 4
2^3  # exponentiation
sqrt(16)  # square root
log(10)   # natural logarithm
exp(1)    # e^1
```
:::

#### Order of Operations

```{r}
#| echo: true
#| eval: false
# Order of operations
2 + 3 * 4
(2 + 3) * 4
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Order of operations
2 + 3 * 4
(2 + 3) * 4
```
:::

#### Mathematical Functions

```{r}
#| echo: true
#| eval: false
# Mathematical functions
abs(-5)
round(3.14159, 2)
ceiling(3.2)
floor(3.8)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Mathematical functions
abs(-5)
round(3.14159, 2)
ceiling(3.2)
floor(3.8)
```
:::

### Variables and Assignment (10 minutes)

Understanding how to store and manipulate data is crucial in R.

#### Creating and Using Variables

```{r}
#| echo: true
#| eval: false
# Creating variables
x <- 10
y <- 5
name <- "Alice"
university <- "State University"

# Using variables
result <- x + y
message <- paste("Hello", name, "from", university)
print(result)
print(message)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Creating variables
x <- 10
y <- 5
name <- "Alice"
university <- "State University"

# Using variables
result <- x + y
message <- paste("Hello", name, "from", university)
print(result)
print(message)
```
:::

#### Multiple Assignment and Naming Conventions

```{r}
#| echo: true
#| eval: false
# Multiple assignment
a <- b <- c <- 100
print(c(a, b, c))

# Variable naming conventions
student_name <- "John Doe"  # snake_case (recommended)
studentName <- "Jane Smith"  # camelCase
student.name <- "Bob Johnson"  # dot notation (less common)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Multiple assignment
a <- b <- c <- 100
print(c(a, b, c))

# Variable naming conventions
student_name <- "John Doe"  # snake_case (recommended)
studentName <- "Jane Smith"  # camelCase
student.name <- "Bob Johnson"  # dot notation (less common)
```
:::

#### Environment Management

```{r}
#| echo: true
#| eval: false
# Check what's in your environment
ls()  # List all variables
rm(x, y)  # Remove specific variables
ls()  # Check again
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Check what's in your environment
ls()  # List all variables
rm(x, y)  # Remove specific variables
ls()  # Check again
```
:::

### Data Types and Structures (15 minutes)

R has several data types. Understanding them is fundamental to effective programming.

#### Basic Data Types

```{r}
#| echo: true
#| eval: false
# Basic data types
age <- 25L          # Integer (note the L)
height <- 5.8       # Numeric (double)
student_name <- "John Doe"  # Character
is_enrolled <- TRUE # Logical
grade <- 'A'        # Character (single quotes also work)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Basic data types
age <- 25L          # Integer (note the L)
height <- 5.8       # Numeric (double)
student_name <- "John Doe"  # Character
is_enrolled <- TRUE # Logical
grade <- 'A'        # Character (single quotes also work)
```
:::

#### Checking Data Types

```{r}
#| echo: true
#| eval: false
# Check data types
typeof(age)
typeof(height)
typeof(student_name)
typeof(is_enrolled)

# Also use class() function
class(age)
class(height)
class(student_name)
class(is_enrolled)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Check data types
typeof(age)
typeof(height)
typeof(student_name)
typeof(is_enrolled)

# Also use class() function
class(age)
class(height)
class(student_name)
class(is_enrolled)
```
:::

#### Type Conversion

```{r}
#| echo: true
#| eval: false
# Type conversion
as.numeric("123")
as.character(456)
as.logical(1)  # 1 becomes TRUE, 0 becomes FALSE
as.integer(3.14)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Type conversion
as.numeric("123")
as.character(456)
as.logical(1)  # 1 becomes TRUE, 0 becomes FALSE
as.integer(3.14)
```
:::

#### Special Values

```{r}
#| echo: true
#| eval: false
# Special values
missing_value <- NA  # Not Available
infinite_value <- Inf
not_a_number <- NaN

# Check for special values
is.na(missing_value)
is.infinite(infinite_value)
is.nan(not_a_number)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Special values
missing_value <- NA  # Not Available
infinite_value <- Inf
not_a_number <- NaN

# Check for special values
is.na(missing_value)
is.infinite(infinite_value)
is.nan(not_a_number)
```
:::

### Vectors: The Building Blocks (15 minutes)

Vectors are the most basic data structure in R. Everything in R is built from vectors.

#### Creating Vectors

```{r}
#| echo: true
#| eval: false
# Creating vectors
grades <- c(85, 92, 78, 96, 88)
student_names <- c("Alice", "Bob", "Charlie", "Diana", "Eve")
passed <- c(TRUE, TRUE, FALSE, TRUE, TRUE)

# Vector properties
length(grades)
names(grades) <- student_names  # Adding names to vector elements
print(grades)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Creating vectors
grades <- c(85, 92, 78, 96, 88)
student_names <- c("Alice", "Bob", "Charlie", "Diana", "Eve")
passed <- c(TRUE, TRUE, FALSE, TRUE, TRUE)

# Vector properties
length(grades)
names(grades) <- student_names  # Adding names to vector elements
print(grades)
```
:::

#### Accessing Vector Elements

```{r}
#| echo: true
#| eval: false
# Accessing vector elements (1-indexed!)
grades[1]              # First element
grades[2:4]            # Elements 2 through 4
grades[c(1, 3, 5)]     # Specific elements
grades["Alice"]        # Using names
grades[grades > 90]    # Conditional indexing
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Accessing vector elements (1-indexed!)
grades[1]              # First element
grades[2:4]            # Elements 2 through 4
grades[c(1, 3, 5)]     # Specific elements
grades["Alice"]        # Using names
grades[grades > 90]    # Conditional indexing
```
:::

#### Vector Operations

```{r}
#| echo: true
#| eval: false
# Vector operations (vectorized!)
grades + 5             # Add 5 to all grades
grades * 1.1           # Multiply all by 1.1
grades > 85            # Logical comparison
which(grades > 85)     # Which positions are TRUE
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Vector operations (vectorized!)
grades + 5             # Add 5 to all grades
grades * 1.1           # Multiply all by 1.1
grades > 85            # Logical comparison
which(grades > 85)     # Which positions are TRUE
```
:::

#### Vector Functions

```{r}
#| echo: true
#| eval: false
# Useful vector functions
sum(grades)
mean(grades)
median(grades)
max(grades)
min(grades)
range(grades)
sort(grades)
sort(grades, decreasing = TRUE)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Useful vector functions
sum(grades)
mean(grades)
median(grades)
max(grades)
min(grades)
range(grades)
sort(grades)
sort(grades, decreasing = TRUE)
```
:::

#### Creating Sequences

```{r}
#| echo: true
#| eval: false
# Creating sequences
seq(1, 10, by = 2)     # 1, 3, 5, 7, 9
seq(0, 1, length.out = 5)  # 5 equally spaced numbers
1:10                   # Shorthand for seq(1, 10, by = 1)
rep(c("A", "B"), times = 3)  # Repeat vector
rep(c("A", "B"), each = 3)   # Repeat each element
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Creating sequences
seq(1, 10, by = 2)     # 1, 3, 5, 7, 9
seq(0, 1, length.out = 5)  # 5 equally spaced numbers
1:10                   # Shorthand for seq(1, 10, by = 1)
rep(c("A", "B"), times = 3)  # Repeat vector
rep(c("A", "B"), each = 3)   # Repeat each element
```
:::

### Working with Data Frames (20 minutes)

Data frames are the primary way to work with structured data in R.

#### Creating a Data Frame

```{r}
#| echo: true
#| eval: false
# Creating a data frame
students <- data.frame(
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  age = c(20, 21, 19, 22, 20),
  grade = c(85, 92, 78, 96, 88),
  major = c("Biology", "Physics", "Chemistry", "Biology", "Math"),
  gpa = c(3.4, 3.7, 3.1, 3.8, 3.5),
  stringsAsFactors = FALSE  # Keep character columns as characters
)

print(students)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Creating a data frame
students <- data.frame(
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  age = c(20, 21, 19, 22, 20),
  grade = c(85, 92, 78, 96, 88),
  major = c("Biology", "Physics", "Chemistry", "Biology", "Math"),
  gpa = c(3.4, 3.7, 3.1, 3.8, 3.5),
  stringsAsFactors = FALSE  # Keep character columns as characters
)

print(students)
```
:::

#### Viewing Data Frame Structure

```{r}
#| echo: true
#| eval: false
# Viewing the data frame
head(students, 3)      # First 3 rows
tail(students, 2)      # Last 2 rows
str(students)          # Structure of the data frame
summary(students)      # Summary statistics
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Viewing the data frame
head(students, 3)      # First 3 rows
tail(students, 2)      # Last 2 rows
str(students)          # Structure of the data frame
summary(students)      # Summary statistics
```
:::

#### Data Frame Properties

```{r}
#| echo: true
#| eval: false
# Data frame dimensions and properties
dim(students)          # Dimensions (rows, columns)
nrow(students)         # Number of rows
ncol(students)         # Number of columns
names(students)        # Column names
rownames(students)     # Row names
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Data frame dimensions and properties
dim(students)          # Dimensions (rows, columns)
nrow(students)         # Number of rows
ncol(students)         # Number of columns
names(students)        # Column names
rownames(students)     # Row names
```
:::

#### Accessing Columns

```{r}
#| echo: true
#| eval: false
# Accessing data frame elements
students$name              # Using $
students[, "grade"]        # Using column name
students[, 3]             # Using column index
students["grade"]         # Returns data frame
students[["grade"]]       # Returns vector
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Accessing data frame elements
students$name              # Using $
students[, "grade"]        # Using column name
students[, 3]             # Using column index
students["grade"]         # Returns data frame
students[["grade"]]       # Returns vector
```
:::

#### Accessing Multiple Columns and Rows

```{r}
#| echo: true
#| eval: false
# Multiple columns
students[, c("name", "grade")]
students[, c(1, 3)]

# Accessing rows
students[1, ]             # First row
students[2:3, ]           # Rows 2 and 3
students[c(1, 3, 5), ]    # Specific rows
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Multiple columns
students[, c("name", "grade")]
students[, c(1, 3)]

# Accessing rows
students[1, ]             # First row
students[2:3, ]           # Rows 2 and 3
students[c(1, 3, 5), ]    # Specific rows
```
:::

#### Accessing Specific Cells and Filtering

```{r}
#| echo: true
#| eval: false
# Accessing specific cells
students[1, "name"]       # First row, name column
students[2, 3]            # Second row, third column

# Conditional filtering
high_performers <- students[students$grade > 90, ]
biology_students <- students[students$major == "Biology", ]
young_high_performers <- students[students$age < 21 & students$grade > 85, ]

print("High performers:")
print(high_performers)
print("Biology students:")
print(biology_students)
print("Young high performers:")
print(young_high_performers)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Accessing specific cells
students[1, "name"]       # First row, name column
students[2, 3]            # Second row, third column

# Conditional filtering
high_performers <- students[students$grade > 90, ]
biology_students <- students[students$major == "Biology", ]
young_high_performers <- students[students$age < 21 & students$grade > 85, ]

print("High performers:")
print(high_performers)
print("Biology students:")
print(biology_students)
print("Young high performers:")
print(young_high_performers)
```
:::

#### Adding and Modifying Columns

```{r}
#| echo: true
#| eval: false
# Adding and modifying columns
students$letter_grade <- ifelse(students$grade >= 90, "A",
                               ifelse(students$grade >= 80, "B",
                               ifelse(students$grade >= 70, "C", "D")))

students$status <- ifelse(students$grade >= 80, "Pass", "Fail")
students$age_group <- cut(students$age, 
                         breaks = c(0, 20, 22, Inf),
                         labels = c("Young", "Medium", "Older"))

print(students)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Adding and modifying columns
students$letter_grade <- ifelse(students$grade >= 90, "A",
                               ifelse(students$grade >= 80, "B",
                               ifelse(students$grade >= 70, "C", "D")))

students$status <- ifelse(students$grade >= 80, "Pass", "Fail")
students$age_group <- cut(students$age, 
                         breaks = c(0, 20, 22, Inf),
                         labels = c("Young", "Medium", "Older"))

print(students)
```
:::

#### Removing and Reordering Columns

```{r}
#| echo: true
#| eval: false
# Removing columns
students$age_group <- NULL  # Remove column

# Reordering columns
students <- students[, c("name", "age", "major", "grade", "gpa", "letter_grade", "status")]

print(students)
```

::: {.callout-tip collapse="true"}
## Show Output

```{r}
#| echo: false
#| eval: true
# Removing columns
students$age_group <- NULL  # Remove column

# Reordering columns
students <- students[, c("name", "age", "major", "grade", "gpa", "letter_grade", "status")]

print(students)
```
:::

------------------------------------------------------------------------

## Practice Exercises

**ðŸ” Exercise 1a:** Create a vector of temperatures in Fahrenheit: 32, 68, 86, 104, 122. Convert them to Celsius using the formula: C = (F - 32) \* 5/9. Find the average temperature in Celsius.

::: {.callout-note collapse="true"}
## Click to see Exercise 1a Answer

```{r}
#| echo: true
#| eval: true
# Create vector of Fahrenheit temperatures
fahrenheit <- c(32, 68, 86, 104, 122)

# Convert to Celsius
celsius <- (fahrenheit - 32) * 5/9

# Display the conversion
print("Fahrenheit temperatures:")
print(fahrenheit)
print("Celsius temperatures:")
print(celsius)

# Find average temperature in Celsius
avg_celsius <- mean(celsius)
print(paste("Average temperature in Celsius:", round(avg_celsius, 2)))
```

The temperatures in Celsius are: 0Â°C, 20Â°C, 30Â°C, 40Â°C, 50Â°C The average temperature in Celsius is 28Â°C.
:::

**ðŸ” Exercise 1b:** Create a data frame with information about 5 books including title, author, year published, and number of pages. Filter for books published after 2000 and calculate the average number of pages.

::: {.callout-note collapse="true"}
## Click to see Exercise 1b Answer

```{r}
#| echo: true
#| eval: true
# Create data frame with book information
books <- data.frame(
  title = c("The Hunger Games", "1984", "To Kill a Mockingbird", 
            "The Fault in Our Stars", "Pride and Prejudice"),
  author = c("Suzanne Collins", "George Orwell", "Harper Lee", 
             "John Green", "Jane Austen"),
  year_published = c(2008, 1949, 1960, 2012, 1813),
  pages = c(374, 328, 281, 313, 432),
  stringsAsFactors = FALSE
)

# Display the original data frame
print("All books:")
print(books)

# Filter for books published after 2000
recent_books <- books[books$year_published > 2000, ]
print("Books published after 2000:")
print(recent_books)

# Calculate average number of pages for recent books
avg_pages_recent <- mean(recent_books$pages)
print(paste("Average number of pages for books published after 2000:", 
            round(avg_pages_recent, 1)))
```

The books published after 2000 are: - The Hunger Games (2008) - 374 pages - The Fault in Our Stars (2012) - 313 pages

The average number of pages for books published after 2000 is 343.5 pages.
:::

------------------------------------------------------------------------
